Teaching (Digital) Epigraphy

### 2015-12-17 source/48/EAGLE16-lame-et-al-teaching.tex
started to update 48
this paper needs a lot of work on tables, images placed in tabular
form, etc.

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Marion Lamé}
\date{2015-12-02}
\begin{document}
Teaching (Digital) Epigraphy

Marion Lamé, Ivan Radman-Livaja, Bruce Robertson and Federico Ponchio

Abstract

Although having the experience of directly manipulating an ancient text-bearing artifact is important for developing the
skills of an epigrapher – an experience that books and photo cannot replace – such access to primary sources is often
problematic. In this article we present our experience with teaching students to transcribe and interpret Roman
inscribed lead tags, using a Digital Autoptic Process (DAP) in a web environment, so to develop basic competences in
epigraphy and digital epigraphy.\footnote{\par \ \  We wish to acknowledge the research assistance of T. M. Crandall,
History student at Mount Allison University \ (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
\ M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
{\textquotedbl}La lettre romaine, entre inscriptions monumentales et écritures communes{\textquotedbl}. The following
students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.}

Keywords

Educational project, digital epigraphy, epigraphy, open access, primary sources

\begin{enumerate}
\item Introduction
\end{enumerate}
Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project \ “Pedagogical,
Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the Tesserarum Sisciae Sylloge (TSS) through
a Digital Autoptic Process (DAP)” we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

2. Epigraphic Primary Sources of Information

@@

\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.752cm]{EAGLE16lameetalteaching-img001.jpg}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.82cm]{EAGLE16lameetalteaching-img002.jpg}
\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like lana, pan(n)um, tvnica,
sagvm, p(a)envla, pal(l)a, palliolum, lodix, banata, and abolla, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\par \ \  Radman-Livaja 2007/11/13/14 and 2013b, 165-172 for analogous tags with Mócsy 1956,
97-104; Egger 1967, 195-210; Frei-Stolba 1984, 127-138; Schwinden 1985 121-137/93, 215-222; Römer-Martijnse 1990/97,
5-48; Feugère 1993, 301-305; Weiss 1991, 211-220; Paci 1995: 29-40; Bassi 1996, 207-216; Bizzarini 2005, 121-135; Buchi
\& Buonopane 2005, 43-51; Cresci \& Pettenò 2010: 42-110; Jacques \& Hoët-Cauwenberghe 2010, 295-317; Wedenig 2013,
237-246.}.

Some of the genuine difficulties of the autoptic process are already solved by a static picture in a print publication.
The tag has a given position, \ it is illuminated from one direction only and therefore it already suggests a reading
direction. A DAP should be able to confront the students to some of those difficulties and to overcome them alone.

@@

3. Digital Epigraphy

In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g., No. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: http://www.labcd.unipi.it/laboratorio .}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\ Terras, 2015
writes about some of the OA issues: “While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.”}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

@@

\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img003.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img004.png}
\end{figure}
Students were involved for a very short period of time: on average two hours per day over a couple of days only (in
Spring 2014 and in Fall 2015). This training was delivered in three steps (“Readings”, “Writings”, “Test \& Taste a
DAP”) during which there were given some academic documentation about the lead tags, and basic knowledge about digital
epigraphy (epigraphic message in a computer, dispositive analysis, existing projects, digital tools, XML and Unicode
encoding, DAP, and issues regarding digital representation of artifacts).

@@

\begin{figure}
\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
\end{figure}
4. Practicing Epigraphy

@@

4.1 Drawing With Help (No. 12563)

@@

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{0.58cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.40900004cm}|m{0.361cm}|m{0.518cm}|m{0.439cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.297cm}|m{0.34900004cm}|}
\hline
\centering M &
\centering V &
\centering R &
\centering T &
\centering I &
\centering O &
\centering L &
\centering A &
\centering M &
\centering P &
\centering V &
\centering I &
\centering\arraybslash S\\\hline
M3 &
V3 &
R7 &
T2 &
I1 &
O3 &
L2 &
A10 &
M2 &
P1 &
V3 &
I1 &
S5\\\hline
M3? &
V4 &
R6 &
T2 &
I1 &
O3 &
L5 &
A10 &
M5 &
P2 &
V4 &
I1 &
S5\\\hline
\end{supertabular}
\end{flushleft}
4.2 Drawing \& Deciphering Without Help (obv. No. 13053)

@@

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.513cm}|m{0.37700003cm}|m{0.518cm}|m{0.361cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.283cm}|m{0.37700003cm}|m{0.46900004cm}|m{0.439cm}|m{0.518cm}|m{0.518cm}|m{0.50200003cm}|m{0.297cm}|m{0.961cm}|}
\hline
\centering 13053 (obv.) &
\centering V &
\centering A &
\centering L &
\centering E &
\centering R &
\centering I &
\centering V &
\centering S &
\centering M &
\centering A &
\centering R &
\centering T &
\centering I &
\centering\arraybslash A\\\hline
\centering Stud. 1 &
V2 &
A10 &
L3 &
E3 &
R7 &
I2 &
V1 &
S6 &
M5 &
? &
R7 &
T3 &
? &
Ignored\\\hline
\centering Stud. 2 &
V2 &
A10 &
L3 &
E2 &
R7 &
I1 &
V1 &
S6? &
M5 &
A10 &
R6? &
T3? &
I1 &
A2\\\hline
\end{supertabular}
\end{flushleft}
\begin{enumerate}
\item \begin{enumerate}
\item Palimpsest (obv. No. 13053)
\end{enumerate}
\end{enumerate}
The older, erased layers, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
g/t relationship between wSystem and tSystem was asked to students.

\begin{figure}
\centering
\includegraphics[width=6.313cm,height=2.207cm]{EAGLE16lameetalteaching-img012.png}
\end{figure}
5. Pedagogical Considerations

@@

6. Conclusion

@@

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Teaching (Digital) Epigraphy}
\author[ilc]{Marion Lamé\corref{first}}
\author[zag]{Ivan Radman-Livaja}
\author[mau]{Bruce Robertson}
\author[isti]{Federico Ponchio}
\address[ilc]{Istituto di Linguistica Computazionale}
\address[isti]{Visual Computing Laboratory, ISTI-CNR}
\address[zag]{Archaeological Museum in Zagreb}
\address[mau]{Mount Allison University}
\cortext[first]{Corresponding author. Email: federico.ponchio@isti.cnr.it}

\date{2015-12-02}
\begin{document}

\maketitle
\begin{abstract}
Although having the experience of directly manipulating an ancient text-bearing artifact is important for developing the
skills of an epigrapher – an experience that books and photo cannot replace – such access to primary sources is often
problematic. In this article we present our experience with teaching students to transcribe and interpret Roman
inscribed lead tags, using a Digital Autoptic Process (DAP) in a web environment, so to develop basic competences in
epigraphy and digital epigraphy.
\end{abstract}
\keywords{Educational project, digital epigraphy, epigraphy, open access, primary sources}

\section{Introduction}

Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the Tesserarum Sisciae Sylloge (TSS) through
a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

\section{Epigraphic Primary Sources of Information}

@@

\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.752cm]{EAGLE16lameetalteaching-img001.jpg}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.82cm]{EAGLE16lameetalteaching-img002.jpg}
\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like lana, pan(n)um, tvnica,
sagvm, p(a)envla, pal(l)a, palliolum, lodix, banata, and abolla, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\citep{radman_livaja_segestica..._2008}, \citep{radman-livaja_roetiquettes_2010}, \citep{radman-livaja_5._2013}, \citep[165-172]{radman-livaja_two_2013}, \citep{radman-livaja_plombs_2014},  for analogous tags with \citep[97-104]{mocsy_olom_1956}; \citep[195-210]{egger_funf_1968}; \citep[127-138]{frei-stolba_les_2011}; \citep[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citep{romer-martijnse_romerzeitliche_1990}; \citep[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citep[301-305]{feugere_etiquette_1993}; \citep[211-220]{weiss_bleietiketten_1991}; \citep[29-40]{paci_etichette_1995}; \citep[207-216]{bassi_tre_1996}; \citep[121-135]{bizzarini_quattro_2005}; \citep[43-51]{buchi_etichette_2005}; \citep[42-110]{cresci_supellex_2010}; \citep[295-317]{jacques_artisanat_2010}; \citep[237-246]{wedenig_bleietikett_2013}.}.

Some of the genuine difficulties of the autoptic process are already solved by a static picture in a print publication.
The tag has a given position, it is illuminated from one direction only and therefore it already suggests a reading
direction. A DAP should be able to confront the students to some of those difficulties and to overcome them alone.

@@

\section{Digital Epigraphy}

In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g., No. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: http://www.labcd.unipi.it/laboratorio .}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\ Terras, 2015
writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

@@

\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img003.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img004.png}
\end{figure}
Students were involved for a very short period of time: on average two hours per day over a couple of days only (in
Spring 2014 and in Fall 2015). This training was delivered in three steps (``Readings'', ``Writings'', ``Test \& Taste a
DAP'') during which there were given some academic documentation about the lead tags, and basic knowledge about digital
epigraphy (epigraphic message in a computer, dispositive analysis, existing projects, digital tools, XML and Unicode
encoding, DAP, and issues regarding digital representation of artifacts).

@@

\begin{figure}
\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
\end{figure}

\section{Practicing Epigraphy}

@@

\subsection{Drawing With Help (No. 12563)}

@@

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{0.58cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.40900004cm}|m{0.361cm}|m{0.518cm}|m{0.439cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.297cm}|m{0.34900004cm}|}
\hline
\centering M &
\centering V &
\centering R &
\centering T &
\centering I &
\centering O &
\centering L &
\centering A &
\centering M &
\centering P &
\centering V &
\centering I &
\centering\arraybslash S\\\hline
M3 &
V3 &
R7 &
T2 &
I1 &
O3 &
L2 &
A10 &
M2 &
P1 &
V3 &
I1 &
S5\\\hline
M3? &
V4 &
R6 &
T2 &
I1 &
O3 &
L5 &
A10 &
M5 &
P2 &
V4 &
I1 &
S5\\\hline
\end{supertabular}
\end{flushleft}

\subsection{Drawing \& Deciphering Without Help (obv. No. 13053)}

@@

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.513cm}|m{0.37700003cm}|m{0.518cm}|m{0.361cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.283cm}|m{0.37700003cm}|m{0.46900004cm}|m{0.439cm}|m{0.518cm}|m{0.518cm}|m{0.50200003cm}|m{0.297cm}|m{0.961cm}|}
\hline
\centering 13053 (obv.) &
\centering V &
\centering A &
\centering L &
\centering E &
\centering R &
\centering I &
\centering V &
\centering S &
\centering M &
\centering A &
\centering R &
\centering T &
\centering I &
\centering\arraybslash A\\\hline
\centering Stud. 1 &
V2 &
A10 &
L3 &
E3 &
R7 &
I2 &
V1 &
S6 &
M5 &
? &
R7 &
T3 &
? &
Ignored\\\hline
\centering Stud. 2 &
V2 &
A10 &
L3 &
E2 &
R7 &
I1 &
V1 &
S6? &
M5 &
A10 &
R6? &
T3? &
I1 &
A2\\\hline
\end{supertabular}
\end{flushleft}

\subsection{Palimpsest (obv. No. 13053)}

The older, erased layers, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
g/t relationship between wSystem and tSystem was asked to students.

\begin{figure}
\centering
\includegraphics[width=6.313cm,height=2.207cm]{EAGLE16lameetalteaching-img012.png}
\end{figure}

\section{Pedagogical Considerations}

@@

\section{Conclusion}

@@

\section*{Aknowledgment}
We wish to acknowledge the research assistance of T. M. Crandall,
History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
``La lettre romaine, entre inscriptions monumentales et écritures communes''. The following
students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.
